---
title: "Geospatial Accuracy in Paleolithic Sites of the Central Mountain Ranges (Northwestern Iberia): A Statistical Study"
author: "Mikel Díaz-Rodríguez"
date: "2024-07-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document contains the code to reproduce the analyses presented in the chapter titled "Geospatial Accuracy in Paleolithic Sites of the Central Mountain Ranges (Northwestern Iberia): A Statistical Study." The procedure is outlined below.

## Import data from csv files

```{r}
sample1 <- read.csv("csv/sample1.csv", header=TRUE, sep=";", stringsAsFactors=F, dec=",")
sample2 <- read.csv("csv/sample2.csv", header=TRUE, sep=";", stringsAsFactors=F, dec=",")
```

## Calculate the difference in distances

```{r, echo=TRUE}
distances <- sample1$distance
```

## Statistical summary

```{r}
summary(distances)
```

## Check normality with Shapiro test

```{r}
shapiro.test(sample1$distance)
```

## Divide the Plots window in 2 parts

```{r}
par(mfrow = c(1, 2))
```

## Create histogram

```{r}
hist(sample1$distance, main = "a. Histogram of non filtered distances",
     xlab = "Distances (m)",
     col = "#fcc5c0")
```

## Create boxplot

```{r}
boxplot(sample1$distance, main = "b. Boxplot of non filtered distances", 
        ylab = "Distances (m)",
        col = "#fcc5c0")
```

## Exclude distance values = 0

```{r}
filtered_sample <- subset(sample1, distance > 0)
```

## Calculate the difference in distances

```{r}
distances2 <- filtered_sample$distance
```

## Statistical summary

```{r}
summary(distances2)
```

## Check normality with Shapiro test

```{r}
shapiro.test(filtered_sample$distance)
```

## Create histogram

```{r}
hist(filtered_sample$distance, main = "a. Histogram of filtered distances", 
     xlab = "Distances (m)",
     col = "#addd8e")
```

## Create boxplot

```{r}
boxplot(filtered_sample$distance, main = "b. Boxplot of filtered distances", 
        ylab = "Distances (m)",
        col = "#addd8e")
```

## Create variables for statistics

```{r}
ALTA1 <- sample1$ALTA
ALTA2 <- sample2$ALTA

HYDROC1 <- sample1$HYDROC
HYDROC2 <- sample2$HYDROC

ASP1 <- sample1$ASP
ASP2 <- sample2$ASP
```

## Statistical summary

```{r}
summary(ALTA1)
summary(ALTA2)

summary(HYDROC1)
summary(HYDROC2)

summary(ASP1)
summary(ASP2)
```

## Add column

Add 'group' column to each sample to identify where the data comes from.

```{r}
sample1$group <- "Not corrected"
sample2$group <- "Corrected"
```

## Combine the samples into a single dataset

```{r}
combined_sample <- rbind(sample1, sample2)
```

## Divide the Plots window in 3 parts

```{r}
par(mfrow = c(1, 3))
```

## ALTA Boxplot

```{r}
boxplot(ALTA ~ group, data = combined_sample, 
        main = "a. ALTA Boxplot", 
        ylab = "ALTA (masl)",
        col = c("#d8daeb", "#fee0b6"), # colours to differentiate the samples
        border = "black")
```

## HYDROC Boxplot

```{r}
boxplot(HYDROC ~ group, data = combined_sample, 
        main = "b. HYDROC boxplot", 
        ylab = "HYDROC (mintues)",
        col = c("#d8daeb", "#fee0b6"), 
        border = "black")
```

## ASP Boxplot

```{r}
boxplot(ASP ~ group, data = combined_sample, 
        main = "c. ASP Boxplot", 
        ylab = "ASP (degrees)",
        col = c("#d8daeb", "#fee0b6"), 
        border = "black")
```

## Statistical tests

```{r}
t.test(ALTA ~ group, data = combined_sample)
wilcox.test(ALTA ~ group, data = combined_sample)

t.test(HYDROC ~ group, data = combined_sample)
wilcox.test(HYDROC ~ group, data = combined_sample)

t.test(ASP ~ group, data = combined_sample)
wilcox.test(ASP ~ group, data = combined_sample)
```

## Exclude sites with distance values = 0

```{r}
sample1 <- sample1[sample1$distance != 0, ]
sample2 <- sample2[sample2$distance != 0, ]
```

## Boxplots to create each variable: ALTA, HYDROC and ASP

```{r}
boxplot(sample1$ALTA, sample2$ALTA, 
        main="a. ALTA boxplot",
        ylab="Altitude (masl)",
        col = c("#80b1d3", "#ffffb3"),
        names=c("Not corrected", "Corrected"))

boxplot(sample1$HYDROC, sample2$HYDROC, 
        main="b. HYDROC boxplot",
        ylab="Cost to potential hydrology (minutes)",
        col = c("#80b1d3", "#ffffb3"),
        names=c("Not corrected", "Corrected"))

boxplot(sample1$ASP, sample2$ASP, 
        main="c. ASP boxplot",
        ylab="Aspect (degrees)",
        col = c("#80b1d3", "#ffffb3"),
        names=c("Not corrected", "Corrected"))
```

## Statistical tests

```{r}
t.test(sample1$ALTA, sample2$ALTA)
wilcox.test(sample1$ALTA, sample2$ALTA)

t.test(sample1$HYDROC, sample2$HYDROC)
wilcox.test(sample1$HYDROC, sample2$HYDROC)

t.test(sample1$ASP, sample2$ASP)
wilcox.test(sample1$ASP, sample2$ASP)
```

## Create variables for statistics

```{r}
ALTA1 <- sample1$ALTA
ALTA2 <- sample2$ALTA

HYDROC1 <- sample1$HYDROC
HYDROC2 <- sample2$HYDROC

ASP1 <- sample1$ASP
ASP2 <- sample2$ASP
```

## Statistical summary

```{r}
summary(ALTA1)
summary(ALTA2)

summary(HYDROC1)
summary(HYDROC2)

summary(ASP1)
summary(ASP2)
```
